---
title: "K-Means and K-Medoids: Runtime Scaling, Cluster Membership, and True Class Shapes"
author: "Halil Bisgin (updated)"
output: html_document
---

## Purpose

This RMarkdown runs K-Means and K-Medoids (PAM) across multiple dataset sizes (n = 300, 500, ..., 1500), records run times, plots runtime curves for both methods, and visualizes final clustering for the largest dataset. Points are drawn with shape corresponding to their true class and color corresponding to the clustering result. Medoids are marked with a black edge.

```{r setup, message=FALSE, warning=FALSE}
# Uncomment to install if needed:
# install.packages(c("cluster", "ggplot2", "tictoc", "dplyr", "tidyr", "patchwork"))

library(cluster)   # pam
library(ggplot2)
library(dplyr)
library(tidyr)
library(patchwork)  # for side-by-side plots
set.seed(42)
```

## Parameters: list of sample sizes to try

```{r params}
# Vector of data set sizes (you asked for 300 to 1500 in increments of 200)
ns <- seq(300, 1500, by = 200)  # 300, 500, 700, ..., 1500

# Number of clusters
k <- 3

# kmeans control
kmeans_nstart <- 10

# store runtime results
runtime_results <- data.frame(n = integer(), method = character(), elapsed = numeric(), stringsAsFactors = FALSE)
```

## Run experiments: simulate data for each n and time K-Means and PAM

```{r run-experiments, message=FALSE}
for (n in ns) {
  # create roughly equal cluster sizes
  n1 <- floor(n/3)
  n2 <- floor(n/3)
  n3 <- n - n1 - n2

  # simulate 3 Gaussian clusters (same centers as before)
  cluster1 <- data.frame(feature1 = rnorm(n1, mean = 2, sd = 0.5),
                         feature2 = rnorm(n1, mean = 2, sd = 0.5),
                         TrueClass = factor(rep(1, n1)))
  cluster2 <- data.frame(feature1 = rnorm(n2, mean = 6, sd = 0.5),
                         feature2 = rnorm(n2, mean = 6, sd = 0.5),
                         TrueClass = factor(rep(2, n2)))
  cluster3 <- data.frame(feature1 = rnorm(n3, mean = 10, sd = 0.5),
                         feature2 = rnorm(n3, mean = 2, sd = 0.5),
                         TrueClass = factor(rep(3, n3)))

  df <- bind_rows(cluster1, cluster2, cluster3)

  # scale features for clustering
  scaled_df <- scale(df[, c("feature1", "feature2")])

  # time K-Means
  t_km <- system.time({
    km_res <- kmeans(scaled_df, centers = k, nstart = kmeans_nstart)
  })
  runtime_results <- runtime_results %>%
    add_row(n = n, method = "KMeans", elapsed = as.numeric(t_km["elapsed"]))

  # time K-Medoids / PAM
  t_kmed <- system.time({
    # pam can be slow for very large n; for these sizes it's usually fine
    pam_res <- pam(scaled_df, k = k)
  })
  runtime_results <- runtime_results %>%
    add_row(n = n, method = "KMedoids", elapsed = as.numeric(t_kmed["elapsed"]))
}
```

## Runtime curves

```{r runtime-plot, fig.width=7, fig.height=4}
runtime_results$method <- factor(runtime_results$method, levels = c("KMeans", "KMedoids"))

ggplot(runtime_results, aes(x = n, y = elapsed, color = method)) +
  geom_point(size = 2) +
  geom_line() +
  scale_color_brewer(palette = "Set1") +
  labs(title = "Runtime vs. number of data points",
       x = "Number of data points (n)",
       y = "Elapsed time (seconds)",
       color = "Method") +
  theme_minimal()
```

## Detailed visualization for the largest dataset (n = max(ns))

This section re-simulates the largest dataset and plots the clustering assignments. Point shape corresponds to the true class label and color corresponds to the cluster membership. For the K-Medoids plot, medoids are marked with black edge color.

```{r large-dataset-visuals, fig.width=12, fig.height=5}
# Simulate one dataset for the largest n (for visualization)
n_large <- max(ns)
n1 <- floor(n_large/3)
n2 <- floor(n_large/3)
n3 <- n_large - n1 - n2

cluster1 <- data.frame(feature1 = rnorm(n1, mean = 2, sd = 0.5),
                       feature2 = rnorm(n1, mean = 2, sd = 0.5),
                       TrueClass = factor(rep(1, n1)))
cluster2 <- data.frame(feature1 = rnorm(n2, mean = 6, sd = 0.5),
                       feature2 = rnorm(n2, mean = 6, sd = 0.5),
                       TrueClass = factor(rep(2, n2)))
cluster3 <- data.frame(feature1 = rnorm(n3, mean = 10, sd = 0.5),
                       feature2 = rnorm(n3, mean = 2, sd = 0.5),
                       TrueClass = factor(rep(3, n3)))

data_large <- bind_rows(cluster1, cluster2, cluster3)

# Scale for clustering and run algorithms
scaled_large <- scale(data_large[, c("feature1", "feature2")])
km_large <- kmeans(scaled_large, centers = k, nstart = kmeans_nstart)
pam_large <- pam(scaled_large, k = k)

# Attach clustering labels
data_large$KMeans_Cluster <- factor(km_large$cluster)
data_large$KMedoids_Cluster <- factor(pam_large$clustering)

# Prepare medoids coordinates (convert back to original scale)
# pam_large$medoids are in scaled space; convert to original scale
medoids_scaled <- pam_large$medoids
medoids_orig <- as.data.frame(sweep(medoids_scaled, 2, attr(scaled_large, "scaled:scale"), "*"))
medoids_orig <- as.data.frame(sweep(medoids_orig, 2, attr(scaled_large, "scaled:center"), "+"))
colnames(medoids_orig) <- c("feature1", "feature2")
# get medoid cluster membership (use clustering assignment of each medoid index)
# pam_large$id.med gives indices of medoids in the original data (if available)
if (!is.null(pam_large$id.med)) {
  medoid_indices <- pam_large$id.med
  medoids_orig$KMedoids_Cluster <- data_large$KMedoids_Cluster[medoid_indices]
} else {
  # fallback: assign cluster number in order (rare)
  medoids_orig$KMedoids_Cluster <- factor(1:k)
}

# Compute kmeans centers in original scale
centers_scaled <- km_large$centers
centers_orig <- as.data.frame(sweep(centers_scaled, 2, attr(scaled_large, "scaled:scale"), "*"))
centers_orig <- as.data.frame(sweep(centers_orig, 2, attr(scaled_large, "scaled:center"), "+"))
colnames(centers_orig) <- c("feature1", "feature2")
centers_orig$KMeans_Cluster <- factor(1:k)

# Plot: K-Means (color = cluster membership, shape = true class)
p_km <- ggplot(data_large, aes(x = feature1, y = feature2)) +
  geom_point(aes(color = KMeans_Cluster, shape = TrueClass), size = 2, alpha = 0.9) +
  scale_color_brewer(palette = "Set1", name = "KMeans\nCluster") +
  scale_shape_manual(values = c(16, 17, 15), name = "True class") +
  geom_point(data = centers_orig, aes(x = feature1, y = feature2),
             color = "black", fill = "yellow", shape = 23, size = 4, stroke = 1) +
  ggtitle(paste0("K-Means clustering (n = ", n_large, ")")) +
  theme_minimal()

# Plot: K-Medoids (color = cluster membership, shape = true class); medoids marked with black edge
p_kmed <- ggplot(data_large, aes(x = feature1, y = feature2)) +
  # main points: color = cluster membership, shape = true class
  geom_point(aes(color = KMedoids_Cluster, shape = TrueClass), size = 2, alpha = 0.9) +
  scale_color_brewer(palette = "Set2", name = "KMedoids\nCluster") +
  scale_shape_manual(values = c(16, 17, 15), name = "True class") +
  # medoids: use shape 21 so we can set a black border (color="black") and fill by cluster
  geom_point(data = medoids_orig,
             aes(x = feature1, y = feature2, fill = KMedoids_Cluster),
             shape = 21, color = "black", size = 4, stroke = 1.2) +
  scale_fill_brewer(palette = "Set2", guide = FALSE) + # fill same palette as color
  ggtitle(paste0("K-Medoids (PAM) clustering (n = ", n_large, ")  — medoids have black edges")) +
  theme_minimal()

# Show side-by-side
p_km + p_kmed
```

## Notes and suggestions

- You can increase the maximum n if your system allows; PAM scales worse than K-Means so PAM runtimes grow faster.
- The runtime plot gives an empirical sense of scaling for these algorithms on this simple 2D data.
- The visualization uses point shape to indicate the true class (shape) and color to indicate cluster membership; medoids are emphasized with black-bordered filled symbols.

```{r sessioninfo}
sessionInfo()
```